---
title: "Visualizations"
author: "Sarah Greenwood"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ggplot workshop
### Setup:
Load libraries and data for the workshop. Tidyverse includes ggplot. 

```{r}
library(tidyverse)
library(gapminder)
library(readr)
data(gapminder)
```

```{r}
schooldata <- read_csv("~/team/bootcamp/R/schooldata.csv")

```

### Base R Histogram
Automatically decides number of bins, labels, and returns counts.  
```{r}
#hist(schooldata$grade)
```

Adding breaks changes the number of bins.
```{r}
hist(schooldata$grade, breaks = 30)
```

```{r}
hist(schooldata$grade, 
     breaks = seq(.875, 4.125, .25), # this controls where the breaks are
     xlim = c(0,4), # lower and upper limits of x axis
     main = "Grade Distribution", # main title of graph
     xlab = "Grade", # x label 
     ylab = "Count") # y label
```

```{r}
plot(x=gapminder$gdpPercap[gapminder$year==2002],
     y=gapminder$lifeExp[gapminder$year==2002], 
     log="x")
pdf("plot1.pdf", width = 6, height = 4)

#return plot
dev.off() 
```

### ggplot 
```{r}
ggplot(schooldata, aes(grade)) +
  geom_histogram(bins = 20)
```

```{r}
ggplot(schooldata, aes(grade)) +
  geom_histogram(binwidth = .25) + 
  xlim(0, NA) + #this tells R to figure out the top bound itself
  xlab("Grade") + # x label 
  ylab("Count") + # y label
  ggtitle("Grade Distribution") # graph title
```

The real power of ggplot comes when we want to split up our data. 

```{r}
ggplot(schooldata, aes(grade)) +
  geom_histogram(binwidth = .25) + 
  xlim(0, NA) +
  xlab("Grade") + 
  ylab("Count") + 
  ggtitle("Grade Distribution") +
  facet_grid(gender ~ .) # formula syntax "I'm going to split by gender..."
```
What if I want to remove the missing data? 

```{r}
schooldata %>% 
  filter(!is.na(gender)) %>% # WAHOO piping is powerful
  mutate(gender = factor(gender, # this changes the order/labels on the male/female part
                         levels = c("M", "F"), 
                         labels = c("Male", "Female"))) %>%
  ggplot(aes(grade)) +
  geom_histogram(binwidth = .25) + 
  xlim(0, NA) +
  xlab("Grade") + 
  ylab("Count") + 
  ggtitle("Grade Distribution") +
  facet_grid(gender ~ .) 
```

#### Scatterplot

```{r}
gapminder %>%
  filter(year==2002) %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) + 
  geom_point() + 
  scale_x_log10() # this transforms the data
```

```{r}
gapminder %>%
  filter(year==2002) %>%
  ggplot(aes(x = gdpPercap, y = lifeExp, color = continent)) + # adds color!
  geom_point() + 
  scale_x_log10() # this transforms the data

ggsave("p1.pdf", width = 6, height = 4) # this saves the most recent plot
```

# Part 2 of Notes

```{r}
library(babynames)
data(babynames)
```

```{r}
babynames <- mutate(babynames, initial = substr(name, 1,1))

initial_props <-
  babynames %>%
  group_by(year, sex, initial) %>%
  summarize(count = sum(n), totalprop = sum(prop)) %>%
  ungroup()

initial_props %>% 
  filter(sex == "F" & initial == "A") %>%
  ggplot(aes(y = totalprop, x = year)) + 
  geom_line()
```

```{r}
initial_props %>% 
  filter(initial == "A") %>%
  ggplot(aes(y = totalprop, x = year, color = sex)) + # this changes the color
  geom_line()
```

```{r}
initial_props %>% 
  filter(initial == "A") %>%
  ggplot(aes(y = totalprop, x = year, linetype = sex)) + # differentiate sex by line type
  geom_line()
```

```{r}
initial_props %>% 
  filter(initial == "A") %>%
  ggplot(aes(y = totalprop, x = year)) + 
  geom_line(aes(linetype = sex), color = "red", size = 1) # general plot characteristics, not mapping from the data
```

```{r}
initial_props %>% 
  ggplot(aes(y = totalprop, x = year, color = sex)) + 
  geom_line() + 
  facet_wrap(~initial) + 
  scale_color_discrete(labels = c("Female", "Male"),
                       name = "Sex") + 
  scale_y_continuous(breaks = c(0, .1, .2)) +
  theme(axis.text.x = element_text(angle=-90, vjust=0.5),
        panel.grid.minor = element_blank()) +  #controls some of the visual elements of the plot 
  labs(title = "Baby Names First Initial", 
       x = "Year", 
       y = "Proportion of Babies")
```


## Unisex Names
The longest command known to mankind (jk but pretty long)

```{r}
# We want a new dataset that's just the top names since 1950

topnames <- babynames %>%
  filter(year >= 1950) %>%
  group_by(name) %>%
  summarize(total = sum(n)) %>%
  arrange(desc(total)) %>%
  slice(1:1000) %>%
  inner_join(babynames) %>% # the inner join gets back the other columns 
  filter(year >= 1950) %>%
  mutate(sex = recode(sex, "F" = "Female", "M" = "Male")) %>%
  group_by(name, sex, total) %>% 
  summarize(sextotal = sum(n)) %>%
  spread(key = sex, value = sextotal) %>%
  mutate(ratio = (Male - Female)/total)

```


```{r}
topnames %>%
  filter(is.na(Male) | is.na(Female))

# figure out which value is throwing the error... no one has ever named a boy Penelope
```

```{r}
ggplot(topnames, aes(x = Male, y = Female, color = ratio)) + 
  geom_point() + 
  lims(x = c(0, 250000), y = c(0,250000)) + 
  geom_abline() + 
  scale_color_gradient(low = "pink", high = "blue", 
                       name = "Sex Dominance", 
                       breaks = c(.9, 0, -.9), 
                       labels = c("Male", "Neutral", "Female")) + 
  geom_text(aes(label = ifelse(abs(ratio) < .2, 
                               name, 
                               '')),
            hjust = -.25, vjust = 0.5, color = "grey10", 
            fontface = "bold", size = 3)
```

Let's do a histogram! We need to format the data in different ways, though. (Wider data -> Longer data)
```{r}
topnames2 <- gather(topnames, key = "sex", value = "sextotal", Male, Female) #changes data format back
```

Barchart
```{r}

topnames2 %>%
  filter(abs(ratio) < .3 & !is.na(ratio)) %>%
  ggplot(aes(x = name, y = sextotal, fill = sex)) +
  geom_bar(stat = "identity") + 
  labs(title = "popular Unisex Names", x = "", y = "Count")
```

Proportion would probably be more telling than count...
```{r}
topnames2 %>%
  filter(abs(ratio) < .3 & !is.na(ratio)) %>%
  ggplot(aes(x = name, y = sextotal, fill = sex)) +
  geom_bar(stat = "identity", position = "fill") + # position makes the proportions show up
  geom_hline(yintercept = 0.5) + # same as line y = x but for historgram
  labs(title = "popular Unisex Names", x = "", y = "Count")
```

```{r}
topnames2 %>%
  filter(abs(ratio) < .3 & !is.na(ratio)) %>%
  ggplot(aes(x = name, y = sextotal, fill = sex)) +
  geom_bar(stat = "identity", position = "dodge") + # makes it not stacked
  labs(title = "popular Unisex Names", x = "", y = "Count")
```

```{r}
library(ggthemes)
```

```{r}
topnames2 %>%
  filter(abs(ratio) < .3 & !is.na(ratio)) %>%
  ggplot(aes(x = name, y = sextotal, fill = sex)) +
  geom_bar(stat = "identity") + # makes it not stacked
  labs(title = "popular Unisex Names", x = "", y = "Count") + 
  theme_minimal() + 
  theme(panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank())
```

## Plotly 

A library that has implementations in R and Python (maybe Matlab and others...). This makes a plot with interactive controls and gives you additional styling options. You can wrap plotly around a ggplot to give it additional, nice functionality. 

```{r}
library(plotly)
```

Prepare the data. 

```{r}
babynames <- mutate(babynames, first2 = substr(name, 1, 2))
anames <- babynames %>% 
  filter(initial == "A" & sex == "F") %>%
  group_by(year, first2) %>%
  summarize(count = sum(n), totalprop = sum(prop)) %>%
  ungroup()
```

Make a plot. With this plot you can't really tell the difference between all of the lines because the colors are so similar. 
```{r}
anames %>% 
  ggplot(aes(y=totalprop, x = year, color = first2)) +
  geom_line()
```

`ggplotly` wraps around ggplot and makes the graph interactive/more useful! 
```{r}
anames %>% 
  ggplot(aes(y=totalprop, x = year, color = first2)) +
  geom_line()
ggplotly()
```


```{r}
plot_ly(anames, x=~year, y=~totalprop, color = ~first2, 
        type = "scatter", mode = "lines")
```

Here's a really bad example of a 3D plot (something is messed up here... )
```{r}
plot_ly(anames, x=~year, z=~totalprop, y=~first2, color = ~first2, 
        line = list(width = 4)) %>%
          add_lines() %>%
          layout(scene=list(xaxis = list(title = "Year"), 
                            zaxis = list(title = "Proportion"), 
                            yaxis = list(title = "Starting Letters")))
```

### Fun Facts

- If you go Help > Cheatsheets in RStudio, you'll find a ton of nifty cheatsheets!
- There's a theme for FiveThirtyEight `theme_fivethirtyeight()`
- You can save plots as objects and then apply different themes to that object
- ggplot doesn't do 3d graphics, but plotly does!